---
title: "Comparing larval size by parental treatment"
author: "Laura H Spencer"
date: "9/6/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Load libraries 

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(car)
library(arm)
```

#### Load data and reformat 

```{r}
larvae.size <- read.csv("../data/2019-08_new-larvae-measurements.csv", header=T, stringsAsFactors = F)
larvae.size[c("Sample", "Population", "Temperature", "pH")] <- lapply(larvae.size[c("Sample", "Population", "Temperature", "pH")], factor)
larvae.size$DateCollected <- as.Date(larvae.size$DateCollected, format = "%m/%d/%y")
```

#### Correct size data 

Scope was not calibrated accurately. Based on measurements of 12 known lengths, multiply lengths by 1.405669

MaxFeret = best estimate of shell width (length running parallel to hinge) 
MinFeret = best estimate of shell heigh (length running perpendicular to hinge) 

```{r}
larvae.size$MaxFeret.cor <- larvae.size$MaxFeret*1.405669 
larvae.size$MinFeret.cor <- larvae.size$MinFeret*1.405669 
```

Inspect size data  

```{r}
summary(larvae.size[c("Length", "Width", "MaxFeret", "MinFeret", "MaxFeret90", "MaxFeret.cor", "MinFeret.cor")]) 
aggregate(MaxFeret.cor ~ Temperature+pH, larvae.size, mean) 
aggregate(MinFeret.cor ~ Temperature+pH, larvae.size, mean)
hist(larvae.size$MaxFeret.cor, breaks = 100)
hist(larvae.size$MinFeret.cor, breaks = 100)
```

#### How many samples do I have size data for, per group? 

The number of larval samples, which were collect from volitional spawn w/n treatments, is unequal among treatments. Unbalanced design (need to account for this in stats) 

Cohort + Temperature + pH 
```{r}
larvae.size %>%
          group_by(Population, Temperature, pH) %>%
          summarise(count= n_distinct(Sample))
```

Just Temperature + pH (all cohorts combined - how i will assess)

```{r}

larvae.size %>%
          group_by(Temperature, pH) %>%
          summarise(count= n_distinct(Sample))
```


### Generate initial plots  

Inspect all larval length data (not yet summarized for each larval group)

- Shell width ~ shell height, color coded by parental treatment 
- Shell width by parental treatment within each cohort 
- Shell height by parental treatment within each cohort 

```{r, echo=FALSE}
print(max.min.all <- ggplot(subset(larvae.size, MinFeret.cor>80), aes(x=MinFeret, y=MaxFeret)) + geom_point(size=.5, aes(color=Temperature:pH)) + theme_bw(base_size = 12) + labs(title="Larval shell size upon release", y=("Max Feret Diameter (~width), (um)"), x="Min Feret Diameter (~height),") + theme(text = element_text(size = 14)) + scale_color_manual(values=c("blue", "forestgreen", "orange", "red")))

print(max.all <- ggplot(larvae.size, aes(x=Population, y=MaxFeret)) + geom_boxplot(aes(color=Temperature:pH)) + theme_bw(base_size = 12) + labs(title="Larval shell width (Max Feret) by cohort & treatment", y=("Max. Feret Diameter (um)"), x="Cohort") + theme(text = element_text(size = 14)) + scale_color_manual(values=c("blue", "forestgreen", "orange", "red")))

print(min.all <- ggplot(larvae.size, aes(x=Population, y=MinFeret)) + geom_boxplot(aes(color=Temperature:pH)) + theme_bw(base_size = 12) + labs(title="Larval shell height (Min Feret) by cohort & treatment", y=("Min. Feret Diameter (um)"), x="Cohort") + theme(text = element_text(size = 14)) + scale_color_manual(values=c("blue", "forestgreen", "orange", "red")))
```

#### Save plots 

```{r, message=FALSE, results=FALSE}
png("../plots/larval-max-min-all.png")
max.min.all
dev.off()
png("../plots/larval-max-feret-all.png")
max.all
dev.off()
png("../plots/larval-min-feret-all.png")
min.all
dev.off()
```

####  Summarize mean shell height and width for each larval sample   

```{r}
larvae.size.mean <- larvae.size %>%
          group_by(Population, Temperature, pH, Sample) %>%
          summarise(MaxFeret.mean=mean(MaxFeret.cor), MinFeret.mean=mean(MinFeret.cor), date=median(DateCollected))
```

#### Generate plots with average length by parental treatment 

```{r, echo=FALSE}
print(min.mean <- ggplot(larvae.size.mean, aes(x=Temperature:pH, y=MaxFeret.mean)) + geom_boxplot() + geom_jitter(width=0.15, size=2, aes(color=Population)) + theme_bw(base_size = 14) + labs(title="Mean shell width, by treatment & cohort", y=("size (um)"), x="Treatment") + theme(text = element_text(size = 16)) + scale_color_manual(values=c("#4daf4a","#377eb8","#984ea3","#e41a1c")))

print(max.mean <- ggplot(larvae.size.mean, aes(x=Temperature:pH, y=MinFeret.mean)) + geom_boxplot() + geom_jitter(width=0.15, size=2, aes(color=Population)) + theme_bw(base_size = 14) + labs(title="Mean shell height, by treatment & cohort", y=("size (um)"), x="Treatment") + theme(text = element_text(size = 16)) + scale_color_manual(values=c("#4daf4a","#377eb8","#984ea3","#e41a1c")))

print(max.mean.cohort <- ggplot(larvae.size.mean, aes(x=Population, y=MaxFeret.mean, color=Temperature:pH)) + geom_point(shape=8, size=2, position = position_jitterdodge(jitter.width = 0.35)) + geom_boxplot() + theme_bw(base_size = 14) + labs(title="Mean shell width, by treatment & cohort", y=("size (um)"), x="Cohort") + theme(text = element_text(size = 16)) + scale_color_manual(values=c("blue", "forestgreen", "orange", "red")))

print(max.mean.date <- ggplot(larvae.size.mean, aes(x=date, y=MaxFeret.mean)) + geom_jitter(width=0.15, size=2, aes(color=Temperature:pH)) + theme_bw(base_size = 14) + labs(title="Mean shell width, by collection date and treatment", y=("size (um)"), x="Date collected") + theme(text = element_text(size = 16)) + scale_color_manual(values=c("blue", "forestgreen", "orange", "red")))
```

#### Save plots 

```{r, message=FALSE, results=FALSE}
png("../plots/larval-max-feret-mean.png")
min.mean
dev.off()
png("../plots/larval-min-feret-mean.png")
max.mean
dev.off()
png("../plots/larval-max-feret-mean-cohort.png")
max.mean.cohort
dev.off()
png("../plots/larval-max-feret-mean-date.png")
max.mean.date
dev.off()
```

## Statistical analyses

Test for length differences among parental temp, pH treats 

NOTE: unbalanced design, so need to use either Type II or Type III SS. See nice explanation here:  [https://www.r-bloggers.com/anova-%E2%80%93-type-iiiiii-ss-explained/](https://www.r-bloggers.com/anova-%E2%80%93-type-iiiiii-ss-explained/)

####  Test Max. Feret Diameter (best estimate for larval width)

Check ANOVA assumptions 

```{r}
hist(larvae.size.mean$MaxFeret.mean)
shapiro.test(larvae.size.mean$MaxFeret.mean)
bartlett.test(MaxFeret.mean ~ pH, data=larvae.size.mean)
bartlett.test(MaxFeret.mean ~ Temperature, data=larvae.size.mean)
```

Test for sign. differences among parental treatments using 

```{r}
anova(lm(MaxFeret.mean ~ Temperature + pH + Temperature:pH, data=larvae.size.mean)) #no interaction. Assess using type II sums of squares, where I look at effects of pH and temperature after controlling for the other factor 
Anova(lm(MaxFeret.mean ~ Temperature*pH, data=larvae.size.mean), type=2) #both sign. 
summary(lm(MaxFeret.mean ~ Temperature+pH, data=larvae.size.mean)) # Fit not great, though 
```


Try mixed effect model, with Population as random effect

```{r}
display(lmer(MaxFeret.mean ~ Temperature*pH + (1|Population), data=larvae.size.mean))
Anova(lmer(MaxFeret.mean ~ Temperature*pH + (1|Population), data=larvae.size.mean), type=2)

display(lmer(MaxFeret.mean ~ Temperature+pH+ (1|Population), data=larvae.size.mean)) 
Anova(lmer(MaxFeret.mean ~ Temperature+pH + (1|Population), data=larvae.size.mean), type=2) #with Pop as random effect and using type II SS, still sign. influence of temp and pH 
```

#### Test min Feret Diameter (best estimate for larval height)

```{r}
hist(larvae.size.mean$MinFeret.mean)
shapiro.test(larvae.size.mean$MinFeret.mean)
bartlett.test(MinFeret.mean ~ pH, data=larvae.size.mean)
bartlett.test(MinFeret.mean ~ Temperature, data=larvae.size.mean)
anova(lm(MinFeret.mean ~ Temperature + pH + Temperature:pH, data=larvae.size.mean)) #no interaction. Assess using type II sums of squares, where I look at effects of pH and temperature after controlling for the other factor 
Anova(lm(MinFeret.mean ~ Temperature*pH, data=larvae.size.mean), type=2) #both sign. 
summary(lm(MinFeret.mean ~ Temperature+pH, data=larvae.size.mean)) # Fit not great, though 

```

Try mixed effect model, with Population as random effect

```{r}
display(lmer(MinFeret.mean ~ Temperature*pH + (1|Population), data=larvae.size.mean))
Anova(lmer(MinFeret.mean ~ Temperature*pH + (1|Population), data=larvae.size.mean), type=2)
display(lmer(MinFeret.mean ~ Temperature+pH+ (1|Population), data=larvae.size.mean)) 
Anova(lmer(MinFeret.mean ~ Temperature+pH + (1|Population), data=larvae.size.mean), type=2) #with Pop as random effect and using type II SS, still sign. influence of temp and pH 

```

#### Test sign. diff among cohorts only 

```{r}
summary(aov(MinFeret.mean ~ Population, data=larvae.size.mean)) #no 
summary(aov(MaxFeret.mean ~ Population, data=larvae.size.mean)) #no 
```

### Check out larval sizes for samples I already sequenced last year

```{r}
quant2018 <-  as.factor(c(41, 26, 32, 10, 48, 77)) #sample numbers for larvae already sequenced (only n=3 per treat)
quant2018 <- larvae.size[larvae.size$Sample %in% quant2018,]
hist(quant2018$MaxFeret)
shapiro.test(log(quant2018$MaxFeret))
ggplot(quant2018, aes(x=Temperature:pH, y=MaxFeret)) + geom_boxplot() + geom_jitter() + theme_bw(base_size = 14) + labs(title="Larval shell  by treatment", y=("size (um)"), x="Treatment") + theme(text = element_text(size = 16)) 
kruskal.test(quant2018$MaxFeret, g = quant2018$pH) #sign. different - 10-low larger.  
```


