---
title: "QuantSeq-diffexp-multivariate"
output: html_document
---
```{r}
list.of.packages <- c("vegan", "dplyr", "factoextra", "FactoMineR", "RColorBrewer", "tibble", "reshape2") #add new libraries here 
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# Load all libraries 
lapply(list.of.packages, FUN = function(X) {
  do.call("require", list(X)) 
})
getwd()
source("references/biostats.R")
```

# Load data

```{r}
# gonad <- read.csv("results/DESeq/pH.counts.csv", row.names = 1) #counts, without gene ID changed for gene name 
gonad <- read.csv("results/DESeq/pH.counts-annot.csv", row.names = 1) #Gene names swapped for trinity gene ID
head(gonad[1:5, 1:5])
pH.p05.names <- as.character(unlist(as.data.frame(read.csv("results/DESeq/pH.p05.contignames.csv")[2])))
key.gonad <- read.csv("data/quantseq_key.csv", header=T, na.strings = NA)[c("filename", "pH", "sex2", "temperature")]


# gonad.trans <- assay(rld) #or if it's already an object from the QuantSeq-DeSeq notebook
for (i in 1:14){
          print(summary(gonad[,i]))
          print(is.numeric(gonad[,i]))
}

larvae <- read.csv("results/DESeq/Larvae.pH.counts-annot.csv", row.names = 1) #Gene names swapped for trinity gene ID
# larvae <- read.csv("results/DESeq/Larvae.pH.counts.csv", row.names = 1) #counts, without gene ID changed for gene name 
pHLarvae.p05.names <- as.character(unlist(as.data.frame(read.csv(file="results/DESeq/pHLarvae.p05.contignames.csv")[2])))
```

How many contigs? 

```{r}
nrow(gonad)
nrow(larvae)
View(head(larvae))
```


```{r}
gonad.trans <- as.matrix(t(gonad))
gonad.pca <- prcomp(gonad.trans, scale=F) #scale=F for variance-covariance matrix
summary(gonad.pca)
pca.eigenval(gonad.pca) #The Proporation of Variance = %variance 
screeplot(gonad.pca, bstick=TRUE)
ordi.monte(gonad.trans, ord="pca", dim=8, perm = 1000, scale = F)
```

PC1, 2, 3 and 4 appear significant! 

```{r}
head(gonad.pca$rotation)
```

```{r}
# Extract the variables (genes) with largest loadings / eigenvalues 
gonad.pc.eigenv <- as.data.frame(gonad.pca$rotation)[c("PC1", "PC2", "PC3", "PC4")]                 # convert to dataframe, keep only first 4 PC
gonad.pc.eigenv <- tibble::rownames_to_column(gonad.pc.eigenv)                                      # make row names a column 

top.1 <- gonad.pc.eigenv %>% arrange(PC1) %>% top_n(5, wt = PC1) %>% select(rowname, PC1)          # For each PC, extract eigenv. w/ biggest abs(loadings)
bottom.1 <- gonad.pc.eigenv %>% arrange(PC1) %>% top_n(-5, wt = PC1) %>% select(rowname, PC1)     
top.1$axis <- "PC1"
bottom.1$axis <- "PC1"
top.2 <- gonad.pc.eigenv %>% arrange(PC2) %>% top_n(5, wt = PC2) %>% select(rowname, PC2)
bottom.2 <- gonad.pc.eigenv %>% arrange(PC2) %>% top_n(-5, wt = PC2) %>% select(rowname, PC2)
top.2$axis <- "PC2"
bottom.2$axis <- "PC2"
top.3 <- gonad.pc.eigenv %>% arrange(PC3) %>% top_n(5, wt = PC3) %>% select(rowname, PC3)
bottom.3 <- gonad.pc.eigenv %>% arrange(PC3) %>% top_n(-5, wt = PC3) %>% select(rowname, PC3)
top.3$axis <- "PC3"
bottom.3$axis <- "PC3"
top.4 <- gonad.pc.eigenv %>% arrange(PC4) %>% top_n(5, wt = PC4) %>% select(rowname, PC4)
bottom.4 <- gonad.pc.eigenv %>% arrange(PC4) %>% top_n(-5, wt = PC4) %>% select(rowname, PC4)
top.4$axis <- "PC4"
bottom.4$axis <- "PC4"
temp <- list(top.1, bottom.1, top.2, bottom.2, top.3, bottom.3, top.4, bottom.4)

for (i in 1:length(temp)){                                                                          # rename columns in all dataframes created above
          colnames(temp[[i]]) <- c("gene", "eigenvalue", "pc")
}
gonad.eigen.top <- do.call(rbind, temp)                                                             # create one dataframe with top eigenvalues 
write.csv(x = gonad.eigen.top, file = "results/DESeq/gonad.top.eigenvectors.csv")                   # write out file 

```


```{r}
# Which variables are both differentially expressed AND sign. contribute to PC1-4? 
gonad.pc.structure <- pca.structure(gonad.pca, gonad.trans, dim=5, cutoff=0)[1:4]
gonad.pc.diff <- as.data.frame(gonad.pc.structure[rownames(gonad.pc.structure) %in% pH.p05.names,])
gonad.pc.diff[1:4] <- lapply(gonad.pc.diff[1:4], function(x) as.numeric(as.character(x)))
gonad.pc.diff <- gonad.pc.diff %>%
    rownames_to_column('gene') %>%
    filter_if(is.numeric, any_vars(abs(.) >= 0)) %>%
    column_to_rownames('gene')
gonad.pc1.genes <- rownames(subset(gonad.pc.diff, PC1!="NA")[1])
gonad.pc2.genes <- rownames(subset(gonad.pc.diff, PC2!="NA")[2])
gonad.pc3.genes <- rownames(subset(gonad.pc.diff, PC3!="NA")[3])
gonad.pc4.genes <- rownames(subset(gonad.pc.diff, PC4!="NA")[4])
gonad.pc.genes <- list(gonad.pc1.genes, gonad.pc2.genes, gonad.pc3.genes, gonad.pc4.genes)
```

```{r}
print(gonad.scores <- gonad.pca$x)

# Check multivariate normality using pc scores 
hist(unlist(as.data.frame(gonad.scores), use.names=FALSE))
shapiro.test(unlist(as.data.frame(gonad.scores), use.names=FALSE))
```

```{r}
ordiplot(gonad.pca, display = "sites")
print(samples.g <- rownames(gonad.scores))
print(key.gonad <- key.gonad[match(samples.g, key.gonad$filename),])
key.gonad$pH.sex <- as.factor(paste(key.gonad$pH, key.gonad$sex2, sep="."))
key.gonad$col <- c(brewer.pal(4, "Set2"))[key.gonad$pH.sex]
plot(gonad.scores, cex=1.6, pch=16, col=key.gonad$col, xlab="PC1 (17.4%)", ylab="PC2 (13.7%)")
```

```{r}
par(oma=c(1,10,1,1))
fviz_screeplot(gonad.pca, addlabels = TRUE, ylim = c(0, 30))
fviz_contrib(gonad.pca, choice = "var", axes = 1, top=10)
fviz_contrib(gonad.pca, choice = "var", axes = 2, top = 10)
fviz_contrib(gonad.pca, choice = "var", axes = 3, top = 10)
fviz_contrib(gonad.pca, choice = "var", axes = 4, top = 10)
fviz_pca_ind(gonad.pca, label="none", habillage=key.gonad$pH.sex)
a <- as.character(fviz_pca_var(gonad.pca, axes = c(1,2), select.var = list(contrib = 10))$data[["name"]])
b <- as.character(fviz_pca_var(gonad.pca, axes = c(1,3), select.var = list(contrib = 10))$data[["name"]])
c <- as.character(fviz_pca_var(gonad.pca, axes = c(1,4), select.var = list(contrib = 10))$data[["name"]])
d <- as.character(fviz_pca_var(gonad.pca, axes = c(2,3), select.var = list(contrib = 10))$data[["name"]])
e <- as.character(fviz_pca_var(gonad.pca, axes = c(2,4), select.var = list(contrib = 10))$data[["name"]])
f <- as.character(fviz_pca_var(gonad.pca, axes = c(3,4), select.var = list(contrib = 10))$data[["name"]])

gonad.top.6 <- unique(c(a, b, c, d, e, f))

as.data.frame(gonad.pc.structure[rownames(gonad.pc.structure) %in% as.character(gonad.top.6),]) #produce table with structure coefficients for the top 6 contributing genes per PC

# Biplots with eigenvectors that are both 1) significant according to PC analysis, and 2) differentially expressed (pvalue<0.05)
for (i in 1:3) {
          print(fviz_pca_biplot(gonad.pca, label="none", invisible="quali", axes = c(i, i+1), select.var= list(name = c(gonad.pc.genes[[i]], gonad.pc.genes[[i+1]])), habillage=key.gonad$pH.sex, pointsize=4, col.var = "gray50") + scale_shape_manual(values=c(rep(16, times=4))) + scale_color_manual(values = c("gray60", "gray40", "lightsteelblue3", "steelblue"))) #+ scale_color_brewer(palette="Paired") + theme_minimal())
}

for (i in 1:2) {
          print(fviz_pca_biplot(gonad.pca, label="none", invisible="quali", axes = c(i, 4), select.var= list(name = c(gonad.pc.genes[[i]], gonad.pc4.genes)),  habillage=key.gonad$pH.sex, pointsize=4, col.var = "gray50") + scale_shape_manual(values=c(rep(16, times=4))) + scale_color_manual(values = c("gray60", "gray40", "lightsteelblue3", "steelblue")))     
}


fviz_pca_biplot(gonad.pca, label="none", invisible="quali", axes = c(1, 3), select.var= list(name = c(gonad.pc1.genes, gonad.pc3.genes)),  habillage=key.gonad$pH.sex, pointsize=4, col.var = "gray50") + scale_shape_manual(values=c(rep(16, times=4))) + scale_color_manual(values = c("gray60", "gray40", "lightsteelblue3", "steelblue"))
```


## Biplots using the top 10 
```{r}

pdf(file="results/PCA/gonad-biplot-for-title.pdf", width=6.5, height=6.5)
fviz_pca_biplot(gonad.pca, label="var", invisible="quali", axes = c(1, 2), select.var= list(contrib=10),  habillage=key.gonad$pH.sex, pointsize=6, col.var = "gray20", repel=TRUE, ggtheme = theme_minimal()) + scale_shape_manual(values=c(rep(16, times=4))) + scale_color_manual(values = c("gray60", "gray30", "lightsteelblue", "steelblue")) + ggtitle(element_blank())  + theme(legend.position="top", legend.title=element_blank(), legend.key.size = unit(2.5,"line"), legend.text=element_text(size=14), plot.title = element_text(hjust = 0.5, colour="gray30", size = 18), axis.title.y = element_text(colour = "gray40"), axis.title.x = element_text(colour = "gray40")) + guides(shape = guide_legend(override.aes = list(size = 8))) + ggtitle(label="PCA Biplots, gonad tissue by pH and sex") 
dev.off()



pdf(file="results/PCA/gonad-biplot-1x3.pdf", width=6.5, height=6.5)
fviz_pca_biplot(gonad.pca, label="var", invisible="quali", axes = c(1, 3), select.var= list(contrib=10),  habillage=key.gonad$pH.sex, pointsize=6, col.var = "gray20", repel=TRUE, ggtheme = theme_minimal()) + scale_shape_manual(values=c(rep(16, times=4))) + scale_color_manual(values = c("gray60", "gray30", "lightsteelblue", "steelblue")) + ggtitle(element_blank())  + theme(legend.position="top", legend.title=element_blank(), legend.key.size = unit(2.5,"line"), legend.text=element_text(size=14), plot.title = element_text(hjust = 0.5, colour="gray30", size = 18), axis.title.y = element_text(colour = "gray40"), axis.title.x = element_text(colour = "gray40")) + guides(shape = guide_legend(override.aes = list(size = 8)))
dev.off()

pdf(file="results/PCA/gonad-biplot-1x4.pdf", width=6.5, height=6.5)
fviz_pca_biplot(gonad.pca, label="var", invisible="quali", axes = c(1, 4), select.var= list(contrib=10),  habillage=key.gonad$pH.sex, pointsize=6, col.var = "gray20", repel=TRUE, ggtheme = theme_minimal()) + scale_shape_manual(values=c(rep(16, times=4))) + scale_color_manual(values = c("gray60", "gray30", "lightsteelblue", "steelblue")) + ggtitle(element_blank())  + theme(legend.position="top", legend.title=element_blank(), legend.key.size = unit(2.5,"line"), legend.text=element_text(size=14), plot.title = element_text(hjust = 0.5, colour="gray30", size = 18), axis.title.y = element_text(colour = "gray40"), axis.title.x = element_text(colour = "gray40")) + guides(shape = guide_legend(override.aes = list(size = 8)))
dev.off()

pdf(file="results/PCA/gonad-biplot-2x3.pdf", width=6.5, height=6.5)
fviz_pca_biplot(gonad.pca, label="var", invisible="quali", axes = c(2, 3), select.var= list(contrib=10),  habillage=key.gonad$pH.sex, pointsize=6, col.var = "gray20", repel=TRUE, ggtheme = theme_minimal()) + scale_shape_manual(values=c(rep(16, times=4))) + scale_color_manual(values = c("gray60", "gray30", "lightsteelblue", "steelblue")) + ggtitle(element_blank())  + theme(legend.position="top", legend.title=element_blank(), legend.key.size = unit(2.5,"line"), legend.text=element_text(size=14), plot.title = element_text(hjust = 0.5, colour="gray30", size = 18), axis.title.y = element_text(colour = "gray40"), axis.title.x = element_text(colour = "gray40")) + guides(shape = guide_legend(override.aes = list(size = 8)))
dev.off()

pdf(file="results/PCA/gonad-biplot-2x4.pdf", width=6.5, height=6.5)
fviz_pca_biplot(gonad.pca, label="var", invisible="quali", axes = c(2, 4), select.var= list(contrib=10),  habillage=key.gonad$pH.sex, pointsize=6, col.var = "gray20", repel=TRUE, ggtheme = theme_minimal()) + scale_shape_manual(values=c(rep(16, times=4))) + scale_color_manual(values = c("gray60", "gray30", "lightsteelblue", "steelblue")) + ggtitle(element_blank())  + theme(legend.position="top", legend.title=element_blank(), legend.key.size = unit(2.5,"line"), legend.text=element_text(size=14), plot.title = element_text(hjust = 0.5, colour="gray30", size = 18), axis.title.y = element_text(colour = "gray40"), axis.title.x = element_text(colour = "gray40")) + guides(shape = guide_legend(override.aes = list(size = 8)))
dev.off()

pdf(file="results/PCA/gonad-biplot-3x4.pdf", width=6.5, height=6.5)
fviz_pca_biplot(gonad.pca, label="var", invisible="quali", axes = c(3, 4), select.var= list(contrib=10),  habillage=key.gonad$pH.sex, pointsize=6, col.var = "gray20", repel=TRUE, ggtheme = theme_minimal()) + scale_shape_manual(values=c(rep(16, times=4))) + scale_color_manual(values = c("gray60", "gray30", "lightsteelblue", "steelblue")) + ggtitle(element_blank())  + theme(legend.position="top", legend.title=element_blank(), legend.key.size = unit(2.5,"line"), legend.text=element_text(size=14), plot.title = element_text(hjust = 0.5, colour="gray30", size = 18), axis.title.y = element_text(colour = "gray40"), axis.title.x = element_text(colour = "gray40")) + guides(shape = guide_legend(override.aes = list(size = 8)))
dev.off()

```


## Make biplots based on variable loadings 
```{r}

# Should merge file I wrote out with annotations for gene names, and write loop for all plots here 

fviz_pca_biplot(gonad.pca, label="none", invisible="quali", axes = c(1, 3), select.var= list(name = c(subset(gonad.eigen.top, pc =="PC2" | pc == "PC4")$gene)),  habillage=key.gonad$pH.sex, pointsize=4, col.var = "gray50") + scale_shape_manual(values=c(rep(16, times=4))) + scale_color_manual(values = c("gray60", "gray40", "lightsteelblue3", "steelblue"))
```



```{r}
17.4+13.7+9.6+9+7.2
```


## Multipanel plot, counts of genes diff expressed and sign. according to PC analysis 

```{r}

pH.p05.names <-rownames(subset(res.pH.Ordered, pvalue < 0.05)) #padj super high across the board - check out contigs with low pvalues instead ... 
par(mfrow=(c(5,5)), mar=c(1,1,4,0), oma=c(5,4,2,2), col="gray25")
for (i in 1:length(pH.p05.names)) {
          plotCounts(dds.pH.DESeq, gene=pH.p05.names[i], intgroup="pH")
}
```


# Repeat the above with larval samples 

Just PC1 and PC2 significant, according to monticarlo test 

```{r}
larvae.trans <- as.matrix(t(larvae))
larvae.pca <- prcomp(larvae.trans, scale=F)
summary(larvae.pca)
pca.eigenval(larvae.pca)
screeplot(larvae.pca, bstick=TRUE)
ordi.monte(larvae.trans, ord="pca", dim=8, perm = 1000, scale = F)

```


```{r}
head(larvae.pca$rotation)
```

```{r}
larvae.annot <- read.csv("results/DESeq/larvae.pH.p05-annot.csv", header=T)
larvae.pc.structure <- pca.structure(larvae.pca, larvae.trans, dim=5, cutoff=0)[1:2]
larvae.pc.diff <- as.data.frame(larvae.pc.structure[rownames(larvae.pc.structure) %in% larvae.annot$Gene,])
larvae.pc.diff[1:2] <- lapply(larvae.pc.diff[1:2], function(x) as.numeric(as.character(x)))
larvae.pc.diff <- larvae.pc.diff %>%
    rownames_to_column('gene') %>%
    filter_if(is.numeric, any_vars(abs(.) >= 0)) %>%
    column_to_rownames('gene')
larvae.pc1.genes <- rownames(subset(larvae.pc.diff, PC1!="NA")[1])
larvae.pc2.genes <- rownames(subset(larvae.pc.diff, PC2!="NA")[2])
```

```{r}
print(larvae.scores <- larvae.pca$x)

# Check multivariate normality using pc scores 
hist(unlist(as.data.frame(larvae.scores), use.names=FALSE))
shapiro.test(unlist(as.data.frame(larvae.scores), use.names=FALSE))
```

```{r}
ordiplot(larvae.pca, display = "sites")
print(samples.l <- rownames(larvae.scores))
key.larvae <- read.csv("data/quantseq_key.csv", header=T, na.strings = NA)[c("filename", "pH")]
print(key.larvae <- key.larvae[match(samples.l, key.larvae$filename),])
key.larvae$col <- c("gray40", "steelblue")[key.larvae$pH]
plot(larvae.scores, cex=2.5, pch=16, col=key.larvae$col, xlab="PC1 (26.8%)", ylab="PC2 (22.3%)")
```

```{r}
fviz_screeplot(larvae.pca, addlabels = TRUE, ylim = c(0, 30))
fviz_contrib(larvae.pca, choice = "var", axes = 1, top = 10)
fviz_contrib(larvae.pca, choice = "var", axes = 2, top = 10)
fviz_pca_ind(larvae.pca, label="none", habillage=key.larvae$pH)
fviz_pca_var(larvae.pca, select.var = list(contrib = 16)) #cool! 
larvae.top.15 <- fviz_pca_var(larvae.pca, select.var = list(contrib = 16))$data
larvae.top.13 <- fviz_pca_var(larvae.pca, select.var = list(contrib = 13))$data


# Biplots with eigenvectors that are both 1) significant according to PC analysis, and 2) differentially expressed (pvalue<0.05)
fviz_pca_biplot(larvae.pca, label="var", axes = c(1,2),  invisible="quali", select.var= list(name = c(larvae.pc1.genes, larvae.pc2.genes)), habillage=key.larvae$pH, pointsize=5, col.var = "gray30", repel=TRUE, ggtheme = theme_minimal()) + scale_shape_manual(values=c(rep(16, times=2))) + scale_color_manual(values = c("gray50", "steelblue")) + ggtitle(label="PCA Biplot, larval gene expression by parental pH")  + theme(legend.position="top", legend.title=element_blank(), plot.title = element_text(hjust = 0.5, colour="gray40"), axis.title.y = element_text(colour = "gray40"), axis.title.x = element_text(colour = "gray40"))


# Biplots with top 14 contriuting eigenvectors (top 6 per +/-)
fviz_pca_biplot(larvae.pca, label="var", axes = c(1,2),  invisible="quali", select.var= list(contrib=13, names=c(larvae.pc1.genes, larvae.pc2.genes)), habillage=key.larvae$pH, pointsize=5, col.var = "gray30", repel=TRUE, ggtheme = theme_minimal()) + scale_shape_manual(values=c(rep(16, times=2))) + scale_color_manual(values = c("gray50", "steelblue")) + ggtitle(label="PCA Biplot, larval gene expression by parental pH")  + theme(legend.position="top", legend.title=element_blank(), plot.title = element_text(hjust = 0.5, colour="gray40"), axis.title.y = element_text(colour = "gray40"), axis.title.x = element_text(colour = "gray40"))
``` 

```{r}
# Extract the variables (genes) with largest loadings / eigenvalues 
larvae.pc.eigenv <- as.data.frame(larvae.pca$rotation)[c("PC1", "PC2")]                 # convert to dataframe, keep only first 2 PC
larvae.pc.eigenv <- tibble::rownames_to_column(larvae.pc.eigenv)                                      # make row names a column 

top.1 <- larvae.pc.eigenv %>% arrange(PC1) %>% top_n(7, wt = PC1) %>% select(rowname, PC1)          # For each PC, extract eigenv. w/ biggest abs(loadings)
bottom.1 <- larvae.pc.eigenv %>% arrange(PC1) %>% top_n(-2, wt = PC1) %>% select(rowname, PC1)    top.1$axis <- "PC1"
bottom.1$axis <- "PC1"
top.2 <- larvae.pc.eigenv %>% arrange(PC2) %>% top_n(2, wt = PC2) %>% select(rowname, PC2)
bottom.2 <- larvae.pc.eigenv %>% arrange(PC2) %>% top_n(-2, wt = PC2) %>% select(rowname, PC2)
top.2$axis <- "PC2"
bottom.2$axis <- "PC2"

temp <- list(top.1, bottom.1, top.2, bottom.2)

for (i in 1:length(temp)){                                                                          # rename columns in all dataframes created above
          colnames(temp[[i]]) <- c("gene", "eigenvalue", "pc")
}
larvae.eigen.top <- do.call(rbind, temp)                                                             # create one dataframe with top eigenvalues 
write.csv(x = larvae.eigen.top, file = "results/DESeq/larvae.top.eigenvectors.csv")                   # write out file 

# dcast(larvae.eigen.top, gene ~ pc, value.var = "eigenvalue")
# diff <- larvae.pc.eigenv[larvae.pc.eigenv$rowname %in% as.character(larvae.annot$Gene),]

```

Much easier, alternative way (but requires hand typing)
```{r}
larvae.eigen.diff <- unique(c(rownames(larvae.top.13), larvae.pc1.genes, larvae.pc2.genes))

as.data.frame(larvae.pc.eigenv[larvae.pc.eigenv$rowname %in% larvae.top.13,])
as.data.frame(larvae.pc.structure[rownames(larvae.pc.structure) %in% larvae.top.16$name,]) #produce table with structure coefficients for the top 16 contributing genes 

```


```{r}

# Biplots with eigenvectors that are either 1) in the top 13 contributors, or 2) differentially expressed (pvalue<0.05). Lots of duplicates 
pdf(file = "results/PCA/larval-biplot.pdf", width = 6.5, height = 6.5)
fviz_pca_biplot(larvae.pca, label="var", axes = c(1,2),  invisible="quali", select.var= list(name = c(larvae.eigen.diff)), habillage=key.larvae$pH, pointsize=6, col.var = "gray20", repel=TRUE, ggtheme = theme_minimal()) + scale_shape_manual(values=c(rep(16, times=2))) + scale_color_manual(values = c("gray50", "steelblue")) + ggtitle(label="PCA Biplot, larval gene expression by parental pH")  + theme(legend.position="top", legend.title=element_blank(), plot.title = element_text(hjust = 0.5, colour="gray30", size = 18), axis.title.y = element_text(colour = "gray40"), axis.title.x = element_text(colour = "gray40"))
dev.off()


# Just samples (points) with sample labels
fviz_pca_biplot(larvae.pca, label="ind", axes = c(1,2),  invisible= c("quali", "var"), select.var= list(name = c(larvae.eigen.top$gene)), habillage=key.larvae$pH, pointsize=6, col.var = "gray30", repel=TRUE, ggtheme = theme_minimal()) + scale_shape_manual(values=c(rep(16, times=2))) + scale_color_manual(values = c("gray50", "steelblue")) + ggtitle(label="PCA Biplot, larval gene expression by parental pH")  + theme(legend.position="top", legend.title=element_blank(), plot.title = element_text(hjust = 0.5, colour="gray40"), axis.title.y = element_text(colour = "gray40"), axis.title.x = element_text(colour = "gray40"))

```



Compare PCA results between gonad and larvae. Any trends? 

```{r}
# Plot genes with sign. PC loadings in gonad against those in larve (take abolute value)
colnames(gonad.pc.structure) <- c("gonadPC1", "gonadPC2", "gonadPC3", "gonadPC4")
colnames(larvae.pc.structure) <- c("larvaePC1", "larvaePC2")
pc.structure <- merge(x=gonad.pc.structure, y=larvae.pc.structure, by.x="row.names", by.y="row.names", all.x=F, all.y=F)
pc.structure[2:7] <- lapply(pc.structure[2:7], function(x) as.numeric(as.character(x)))

plot(x=abs(pc.structure$gonadPC1), y=abs(pc.structure$larvaePC1), main="gonad PC1 x larvae PC1")
plot(x=abs(pc.structure$gonadPC1), y=abs(pc.structure$larvaePC2), main="gonad PC1 x larvae PC2")
plot(x=abs(pc.structure$gonadPC2), y=abs(pc.structure$larvaePC1), main="gonad PC2 x larvae PC1")
plot(x=abs(pc.structure$gonadPC2), y=abs(pc.structure$larvaePC2), main="gonad PC2 x larvae PC2")
plot(x=abs(pc.structure$gonadPC3), y=abs(pc.structure$larvaePC1), main="gonad PC3 x larvae PC1")
plot(x=abs(pc.structure$gonadPC3), y=abs(pc.structure$larvaePC2), main="gonad PC3 x larvae PC2")
plot(x=abs(pc.structure$gonadPC4), y=abs(pc.structure$larvaePC1), main="gonad PC4 x larvae PC1")
plot(x=abs(pc.structure$gonadPC4), y=abs(pc.structure$larvaePC2), main="gonad PC4 x larvae PC2")

```

# perMANOVA 

```{r}
# reload in the count data (untransformed)
gonad.counts <- read.csv(file="results/multivariate/gonad.counts.csv", header=T, row.names = 1)
gonad.counts.t <- log(gonad.counts+1)
gonad.d<-vegdist(gonad.counts.t,"bray")

larvae.counts <- read.csv(file="results/multivariate/larvae.counts.csv", header=T, row.names=1)
larvae.counts.t <- log(larvae.counts+1)
larvae.d<-vegdist(larvae.counts.t,"bray")
perm.grps <- read.csv("data/quantseq_key.csv", header=T, na.strings = NA, row.names=1)[c("pH", "sex2", "tissue", "temperature")]
perm.grps[perm.grps$tissue=="larvae",]

# permANOVA and dispersion test, gonad by pH
print(gonad.perm.ph <- adonis(gonad.counts.t ~ pH, data=subset(perm.grps, tissue=="gonad") ,permutations=1000,method='bray')) #no global pH diff. 
print(gonad.bdp.ph<-betadisper(gonad.d,subset(perm.grps, tissue=="gonad")$pH))
anova(gonad.bdp.ph) #no difference in dispersion 

# permANOVA and dispersion test, gonad by sex
print(gonad.perm.sex <- adonis(gonad.counts.t ~ sex2, data=subset(perm.grps, tissue=="gonad") ,permutations=1000,method='bray')) #no global sex diff. 
print(gonad.bdp.sex<-betadisper(gonad.d,subset(perm.grps, tissue=="gonad")$sex2))
anova(gonad.bdp.sex) #no difference in dispersion 

# permANOVA and dispersion test, gonad by sex
print(gonad.perm.phsex <- adonis(gonad.counts.t ~ pH + sex2, data=subset(perm.grps, tissue=="gonad") ,permutations=1000,method='bray')) #no global sex+pH diff. 

# permANOVA and dispersion test, gonad by sex
print(gonad.perm.temp <- adonis(gonad.counts.t ~ temperature, data=subset(perm.grps, tissue=="gonad") ,permutations=1000,method='bray')) #no global temp diff. 
print(gonad.bdp.temp<-betadisper(gonad.d,subset(perm.grps, tissue=="gonad")$temperature))
anova(gonad.bdp.temp) #no difference in dispersion 

# permANOVA and dispersion test, larvae by parental pH
print(larvae.perm.ph <- adonis(larvae.counts.t ~ pH, data=subset(perm.grps, tissue=="larvae") ,permutations=1000,method='bray')) #no global pH diff. 
print(larvae.bdp.ph<-betadisper(larvae.d,subset(perm.grps, tissue=="larvae")$pH))
anova(larvae.bdp.ph) #no difference in dispersion 

larvae.d
```



